---
title: "Statistical Inference: Course Project"
output: pdf_document
---

```{r echo=FALSE}
  # Set working directory
  setwd("~/datasciencecoursera/StatsProj1/")
  library(ggplot2)
```
## 1. Show where the distribution is centered at and compare it to the theoretical center of the distribution.

We can visualize the distribution with the histogram:
```{r}
set.seed(5678)
n = 1000; lambda = 0.2;
hist(rexp(n, lambda))
```

n 
```{r}
par(mfrow=c(1,2))
mns = NULL
for (i in 1 : 40) mns = c(mns, mean(rexp(n, lambda)))
hist(mns)
cltmean = cumsum(mns) / (1 : 40)
abline(v=cltmean, col="red", lwd=2)
g <- ggplot(data.frame(x = 1 : 40, y = mns), aes(x = x, y = y))
g <- g + geom_hline(yintercept = 0) + geom_line(size = 2)
g <- g + labs(x = "Number of obs", y = "Cumulative mean")
g
```
The expected value of an exponentially distributed random variable $X$ with rate parameter $\lambda$ is given by $E[X] = \frac{1}{\lambda} = \frac{1}{20} \approx `r 1/20`$. We run a Monte Carlo simulation

we can visually see this is approximately normal:
```{r}
hist(mns)
cltmean = cumsum(mns) / (1 : 40)
abline(v=cltmean, col="red", lwd=2)
```
## 2. Show how variable it is and compare it to the theoretical variance of the distribution.

The theretical variance of the distribution is $Var[X] = \frac{1}{\lambda^2} = \frac{1}{20^2} \approx `r 1/20^2`$. We run a Monte Carlo simulation

## 3. Show that the distribution is approximately normal.

